### 2025-09-18 paper/sec/introduction.tex
Progress on intro.
Progress on introduction.

Recent work on data provenance and data visualisation~\citep{psallidas18smoke,bond25}, on the other hand,
takes a more infrastructural approach to the transparency problem, linking computed outputs to their data
sources directly. This allows visual outputs to support \emph{provenance queries}, user interactions (e.g.
mousing over a visual element) that reveal how visual features relate to data. The advantage of this approach
is that the interactions are provided automatically via trusted infrastructure, namely a query language or
general-purpose programming language which tracks dependencies as data flows through a computation. However,
these approaches are limited to outputs computed from data, such as visualisations. What is missing is a way
to extend these ``direct interrogation'' features to natural language itself, where the main claims of most
scholarly articles are actually made.

In this paper, we this gap by combining these two approaches, leveraging the ability of LLMs to understand
technical language and synthesise queries over data, and utilising the provenance-tracking infrastructure of
an open source programming language called Fluid (\url{https://f.luid.org/}) designed originally for
transparent visualisations~\citep{perera22,bond25}. We show how these two technologies can together support
the creation and deployment of \emph{transparent documents}, web-based scholarly articles where key
quantitative claims expressed in natural language --- such as the claim that one software system is faster
than another at solving a problem --- are computed from the relevant data, and that support interactive
provenance queries that reveal to a reader or reviewer the specific data elements that support the claim.

@@

\paragraph{Contributions}

Our specific contributions are:

>>>>

Recent work on data provenance and data visualisation~\citep{psallidas18smoke,bond25}, on the other hand,
takes a more infrastructural approach to the transparency problem. These approaches link computed outputs to
their data sources directly, by tracking dependency information. This allows visual outputs to support
\emph{provenance queries}, user interactions (e.g. mousing over a visual element) that reveal how visual
features relate to data. The advantage of this approach is that the interactions are provided automatically
via trusted infrastructure, namely a query language or general-purpose programming language which tracks how
data flows through a computation. However, these approaches are limited to outputs computed from data, such as
visualisations. What is missing is a way to extend these ``direct interrogation'' features to natural language
itself, where the main claims of most scholarly articles are actually made.

In this paper, we address this gap by combining these two approaches, leveraging the ability of LLMs to
understand technical language and synthesise queries over data, plus the provenance-tracking infrastructure of
an open source programming language called Fluid (\url{https://f.luid.org/}) designed originally for
transparent visualisations~\citep{perera22,bond25}. We show how these two technologies can together support
the creation and deployment of \emph{transparent documents}, web-based scholarly articles where key
quantitative claims expressed in natural language --- such as the claim that one software system is faster
than another at solving a problem --- are computed from the relevant data, and that support interactive
provenance queries that reveal to a reader or reviewer the specific data elements that support the claim.

@@

\paragraph{Contributions.} Our specific contributions are:
